### 2024-07-29 paper/paper.tex
Auto format paper

\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2024

% ready for submission
\usepackage{neurips_2024}

% to compile a preprint version, e.g., for submission to arXiv, add add the
% [preprint] option:
%     \usepackage[preprint]{neurips_2024}

% to compile a camera-ready version, add the [final] option, e.g.:
%     \usepackage[final]{neurips_2024}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2024}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

\title{Formatting Instructions For NeurIPS 2024}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  David S.~Hippocampus\thanks{Use footnote for providing further information
    about author (webpage, alternative address)---\emph{not} for acknowledging
    funding agencies.} \\
  Department of Computer Science\\
  Cranberry-Lemon University\\
  Pittsburgh, PA 15213 \\
  \texttt{hippo@cs.cranberry-lemon.edu} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}

\maketitle

\begin{abstract}
  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
  both the left- and right-hand margins. Use 10~point type, with a vertical
  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
  bold, and in point size 12. Two line spaces precede the abstract. The abstract
  must be limited to one paragraph.
\end{abstract}

\section{Submission of papers to NeurIPS 2024}

Please read the instructions below carefully and follow them faithfully.

\subsection{Style}

Papers to be submitted to NeurIPS 2024 must be prepared according to the
instructions presented here. Papers may only be up to {\bf nine} pages long,
including figures. Additional pages \emph{containing only acknowledgments and
  references} are allowed. Papers that exceed the page limit will not be
reviewed, or in any other way considered for presentation at the conference.

The margins in 2024 are the same as those in previous years.

Authors are required to use the NeurIPS \LaTeX{} style files obtainable at the
NeurIPS website as indicated below. Please make sure you use the current files
and not previous versions. Tweaking the style files may be grounds for
rejection.

\subsection{Retrieval of style files}

The style files for NeurIPS and other conference information are available on
the website at
\begin{center}
  \url{http://www.neurips.cc/}
\end{center}
The file \verb+neurips_2024.pdf+ contains these instructions and illustrates the
various formatting requirements your NeurIPS paper must satisfy.

The only supported style file for NeurIPS 2024 is \verb+neurips_2024.sty+,
rewritten for \LaTeXe{}.  \textbf{Previous style files for \LaTeX{} 2.09,
  Microsoft Word, and RTF are no longer supported!}

The \LaTeX{} style file contains three optional arguments: \verb+final+, which
creates a camera-ready copy, \verb+preprint+, which creates a preprint for
submission to, e.g., arXiv, and \verb+nonatbib+, which will not load the
\verb+natbib+ package for you in case of package clash.

\paragraph{Preprint option}
If you wish to post a preprint of your work online, e.g., on arXiv, using the
NeurIPS style, please use the \verb+preprint+ option. This will create a
nonanonymized version of your work with the text ``Preprint. Work in progress.''
in the footer. This version may be distributed as you see fit, as long as you do not say which conference it was submitted to. Please \textbf{do
  not} use the \verb+final+ option, which should \textbf{only} be used for
papers accepted to NeurIPS.

At submission time, please omit the \verb+final+ and \verb+preprint+
options. This will anonymize your submission and add line numbers to aid
review. Please do \emph{not} refer to these line numbers in your paper as they
will be removed during generation of camera-ready copies.

The file \verb+neurips_2024.tex+ may be used as a ``shell'' for writing your
paper. All you have to do is replace the author, title, abstract, and text of
the paper with your own.

The formatting instructions contained in these style files are summarized in
Sections \ref{gen_inst}, \ref{headings}, and \ref{others} below.

\section{General formatting instructions}
\label{gen_inst}

The text must be confined within a rectangle 5.5~inches (33~picas) wide and
9~inches (54~picas) long. The left margin is 1.5~inch (9~picas).  Use 10~point
type with a vertical spacing (leading) of 11~points.  Times New Roman is the
preferred typeface throughout, and will be selected for you by default.
Paragraphs are separated by \nicefrac{1}{2}~line space (5.5 points), with no
indentation.

The paper title should be 17~point, initial caps/lower case, bold, centered
between two horizontal rules. The top rule should be 4~points thick and the
bottom rule should be 1~point thick. Allow \nicefrac{1}{4}~inch space above and
below the title to rules. All pages should start at 1~inch (6~picas) from the
top of the page.

For the final version, authors' names are set in boldface, and each name is
centered above the corresponding address. The lead author's name is to be listed
first (left-most), and the co-authors' names (if different address) are set to
follow. If there is only one co-author, list both author and co-author side by
side.

Please pay special attention to the instructions in Section \ref{others}
regarding figures, tables, acknowledgments, and references.

\section{Headings: first level}
\label{headings}

All headings should be lower case (except for first word and proper nouns),
flush left, and bold.

First-level headings should be in 12-point type.

\subsection{Headings: second level}

Second-level headings should be in 10-point type.

\subsubsection{Headings: third level}

Third-level headings should be in 10-point type.

\paragraph{Paragraphs}

There is also a \verb+\paragraph+ command available, which sets the heading in
bold, flush left, and inline with the text, with the heading followed by 1\,em
of space.

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone.

\subsection{Citations within the text}

The \verb+natbib+ package will be loaded for you by default.  Citations may be
author/year or numeric, as long as you maintain internal consistency.  As to the
format of the references themselves, any style is acceptable as long as it is
used consistently.

The documentation for \verb+natbib+ may be found at
\begin{center}
  \url{http://mirrors.ctan.org/macros/latex/contrib/natbib/natnotes.pdf}
\end{center}
Of note is the command \verb+\citet+, which produces citations appropriate for
use in inline text.  For example,
\begin{verbatim}
   \citet{hasselmo} investigated\dots
\end{verbatim}
produces
\begin{quote}
  Hasselmo, et al.\ (1995) investigated\dots
\end{quote}

If you wish to load the \verb+natbib+ package with options, you may add the
following before loading the \verb+neurips_2024+ package:
\begin{verbatim}
   \PassOptionsToPackage{options}{natbib}
\end{verbatim}

If \verb+natbib+ clashes with another package you load, you can add the optional
argument \verb+nonatbib+ when loading the style file:
\begin{verbatim}
   \usepackage[nonatbib]{neurips_2024}
\end{verbatim}

As submission is double blind, refer to your own published work in the third
person. That is, use ``In the previous work of Jones et al.\ [4],'' not ``In our
previous work [4].'' If you cite your other papers that are not widely available
(e.g., a journal paper under review), use anonymous author names in the
citation, e.g., an author of the form ``A.\ Anonymous'' and include a copy of the anonymized paper in the supplementary material.

\subsection{Footnotes}

Footnotes should be used sparingly.  If you do require a footnote, indicate
footnotes with a number\footnote{Sample of the first footnote.} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches (12~picas).

Note that footnotes are properly typeset \emph{after} punctuation
marks.\footnote{As in this example.}

\subsection{Figures}

\begin{figure}
  \centering
  \fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
  \caption{Sample figure caption.}
\end{figure}

All artwork must be neat, clean, and legible. Lines should be dark enough for
purposes of reproduction. The figure number and caption always appear after the
figure. Place one line space before the figure caption and one line space after
the figure. The figure caption should be lower case (except for first word and
proper nouns); figures are numbered consecutively.

You may use color figures.  However, it is best for the figure captions and the
paper body to be legible if the paper is printed in either black/white or in
color.

\subsection{Tables}

All tables must be centered, neat, clean and legible.  The table number and
title always appear before the table.  See Table~\ref{sample-table}.

Place one line space before the table title, one line space after the
table title, and one line space after the table. The table title must
be lower case (except for first word and proper nouns); tables are
numbered consecutively.

Note that publication-quality tables \emph{do not contain vertical rules.} We
strongly suggest the use of the \verb+booktabs+ package, which allows for
typesetting high-quality, professional tables:
\begin{center}
  \url{https://www.ctan.org/pkg/booktabs}
\end{center}
This package was used to typeset Table~\ref{sample-table}.

\begin{table}
  \caption{Sample table title}
  \label{sample-table}
  \centering
  \begin{tabular}{lll}
    \toprule
    \multicolumn{2}{c}{Part}                   \\
    \cmidrule(r){1-2}
    Name     & Description     & Size ($\mu$m) \\
    \midrule
    Dendrite & Input terminal  & $\sim$100     \\
    Axon     & Output terminal & $\sim$10      \\
    Soma     & Cell body       & up to $10^6$  \\
    \bottomrule
  \end{tabular}
\end{table}

\subsection{Math}
Note that display math in bare TeX commands will not create correct line numbers for submission. Please use LaTeX (or AMSTeX) commands for unnumbered display math. (You really shouldn't be using \$\$ anyway; see \url{https://tex.stackexchange.com/questions/503/why-is-preferable-to} and \url{https://tex.stackexchange.com/questions/40492/what-are-the-differences-between-align-equation-and-displaymath} for more information.)

\subsection{Final instructions}

Do not change any aspects of the formatting parameters in the style files.  In
particular, do not modify the width or length of the rectangle the text should
fit into, and do not change font sizes (except perhaps in the
\textbf{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PDF files}

Please prepare submission files with paper size ``US Letter,'' and not, for
example, ``A4.''

Fonts were the main cause of problems in the past years. Your PDF file must only
contain Type 1 or Embedded TrueType fonts. Here are a few instructions to
achieve this.

\begin{itemize}

  \item You should directly generate PDF files using \verb+pdflatex+.

  \item You can check which fonts a PDF files uses.  In Acrobat Reader, select the
        menu Files$>$Document Properties$>$Fonts and select Show All Fonts. You can
        also use the program \verb+pdffonts+ which comes with \verb+xpdf+ and is
        available out-of-the-box on most Linux machines.

  \item \verb+xfig+ "patterned" shapes are implemented with bitmap fonts.  Use
        "solid" shapes instead.

  \item The \verb+\bbold+ package almost always uses bitmap fonts.  You should use
        the equivalent AMS Fonts:
        \begin{verbatim}
   \usepackage{amsfonts}
\end{verbatim}
        followed by, e.g., \verb+\mathbb{R}+, \verb+\mathbb{N}+, or \verb+\mathbb{C}+
        for $\mathbb{R}$, $\mathbb{N}$ or $\mathbb{C}$.  You can also use the following
        workaround for reals, natural and complex:
        \begin{verbatim}
   \newcommand{\RR}{I\!\!R} %real numbers
   \newcommand{\Nat}{I\!\!N} %natural numbers
   \newcommand{\CC}{I\!\!\!\!C} %complex numbers
\end{verbatim}
        Note that \verb+amsfonts+ is automatically loaded by the \verb+amssymb+ package.

\end{itemize}

If your file contains type 3 fonts or non embedded TrueType fonts, we will ask
you to fix it.

\subsection{Margins in \LaTeX{}}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+ from the \verb+graphicx+ package. Always specify the
figure width as a multiple of the line width as in the example below:
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
See Section 4.4 in the graphics bundle documentation
(\url{http://mirrors.ctan.org/macros/latex/required/graphics/grfguide.pdf})

A number of width problems arise when \LaTeX{} cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command when
necessary.

\begin{ack}
  Use unnumbered first level headings for the acknowledgments. All acknowledgments
  go at the end of the paper before the list of references. Moreover, you are required to declare
  funding (financial activities supporting the submitted work) and competing interests (related financial activities outside the submitted work).
  More information about this disclosure can be found at: \url{https://neurips.cc/Conferences/2024/PaperInformation/FundingDisclosure}.

  Do {\bf not} include this section in the anonymized submission, only in the final paper. You can use the \texttt{ack} environment provided in the style file to automatically hide this section in the anonymized submission.
\end{ack}

\section*{References}

References follow the acknowledgments in the camera-ready paper. Use unnumbered first-level heading for
the references. Any choice of citation style is acceptable as long as you are
consistent. It is permissible to reduce the font size to \verb+small+ (9 point)
when listing the references.
Note that the Reference section does not count towards the page limit.
\medskip

{
\small

[1] Alexander, J.A.\ \& Mozer, M.C.\ (1995) Template-based algorithms for
connectionist rule extraction. In G.\ Tesauro, D.S.\ Touretzky and T.K.\ Leen
(eds.), {\it Advances in Neural Information Processing Systems 7},
pp.\ 609--616. Cambridge, MA: MIT Press.

  [2] Bower, J.M.\ \& Beeman, D.\ (1995) {\it The Book of GENESIS: Exploring
    Realistic Neural Models with the GEneral NEural SImulation System.}  New York:
TELOS/Springer--Verlag.

[3] Hasselmo, M.E., Schnell, E.\ \& Barkai, E.\ (1995) Dynamics of learning and
recall at excitatory recurrent synapses and cholinergic modulation in rat
hippocampal region CA3. {\it Journal of Neuroscience} {\bf 15}(7):5249-5262.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\appendix

\section{Appendix / supplemental material}

Optionally include supplemental material (complete proofs, additional experiments and plots) in appendix.
All such materials \textbf{SHOULD be included in the main submission.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{NeurIPS Paper Checklist}

%%% BEGIN INSTRUCTIONS %%%
The checklist is designed to encourage best practices for responsible machine learning research, addressing issues of reproducibility, transparency, research ethics, and societal impact. Do not remove the checklist: {\bf The papers not including the checklist will be desk rejected.} The checklist should follow the references and follow the (optional) supplemental material.  The checklist does NOT count towards the page
limit.

Please read the checklist guidelines carefully for information on how to answer these questions. For each question in the checklist:
\begin{itemize}
  \item You should answer \answerYes{}, \answerNo{}, or \answerNA{}.
  \item \answerNA{} means either that the question is Not Applicable for that particular paper or the relevant information is Not Available.
  \item Please provide a short (1–2 sentence) justification right after your answer (even for NA).
        % \item {\bf The papers not including the checklist will be desk rejected.}
\end{itemize}

{\bf The checklist answers are an integral part of your paper submission.} They are visible to the reviewers, area chairs, senior area chairs, and ethics reviewers. You will be asked to also include it (after eventual revisions) with the final version of your paper, and its final version will be published with the paper.

The reviewers of your paper will be asked to use the checklist as one of the factors in their evaluation. While "\answerYes{}" is generally preferable to "\answerNo{}", it is perfectly acceptable to answer "\answerNo{}" provided a proper justification is given (e.g., "error bars are not reported because it would be too computationally expensive" or "we were unable to find the license for the dataset we used"). In general, answering "\answerNo{}" or "\answerNA{}" is not grounds for rejection. While the questions are phrased in a binary way, we acknowledge that the true answer is often more nuanced, so please just use your best judgment and write a justification to elaborate. All supporting evidence can appear either in the main paper or the supplemental material, provided in appendix. If you answer \answerYes{} to a question, in the justification please point to the section(s) where related material for the question can be found.

IMPORTANT, please:
\begin{itemize}
  \item {\bf Delete this instruction block, but keep the section heading ``NeurIPS paper checklist"},
  \item  {\bf Keep the checklist subsection headings, questions/answers and guidelines below.}
  \item {\bf Do not modify the questions and only use the provided macros for your answers}.
\end{itemize}

%%% END INSTRUCTIONS %%%

\begin{enumerate}

  \item {\bf Claims}
  \item[] Question: Do the main claims made in the abstract and introduction accurately reflect the paper's contributions and scope?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the abstract and introduction do not include the claims made in the paper.
          \item The abstract and/or introduction should clearly state the claims made, including the contributions made in the paper and important assumptions and limitations. A No or NA answer to this question will not be perceived well by the reviewers.
          \item The claims made should match theoretical and experimental results, and reflect how much the results can be expected to generalize to other settings.
          \item It is fine to include aspirational goals as motivation as long as it is clear that these goals are not attained by the paper.
        \end{itemize}

  \item {\bf Limitations}
  \item[] Question: Does the paper discuss the limitations of the work performed by the authors?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper has no limitation while the answer No means that the paper has limitations, but those are not discussed in the paper.
          \item The authors are encouraged to create a separate "Limitations" section in their paper.
          \item The paper should point out any strong assumptions and how robust the results are to violations of these assumptions (e.g., independence assumptions, noiseless settings, model well-specification, asymptotic approximations only holding locally). The authors should reflect on how these assumptions might be violated in practice and what the implications would be.
          \item The authors should reflect on the scope of the claims made, e.g., if the approach was only tested on a few datasets or with a few runs. In general, empirical results often depend on implicit assumptions, which should be articulated.
          \item The authors should reflect on the factors that influence the performance of the approach. For example, a facial recognition algorithm may perform poorly when image resolution is low or images are taken in low lighting. Or a speech-to-text system might not be used reliably to provide closed captions for online lectures because it fails to handle technical jargon.
          \item The authors should discuss the computational efficiency of the proposed algorithms and how they scale with dataset size.
          \item If applicable, the authors should discuss possible limitations of their approach to address problems of privacy and fairness.
          \item While the authors might fear that complete honesty about limitations might be used by reviewers as grounds for rejection, a worse outcome might be that reviewers discover limitations that aren't acknowledged in the paper. The authors should use their best judgment and recognize that individual actions in favor of transparency play an important role in developing norms that preserve the integrity of the community. Reviewers will be specifically instructed to not penalize honesty concerning limitations.
        \end{itemize}

  \item {\bf Theory Assumptions and Proofs}
  \item[] Question: For each theoretical result, does the paper provide the full set of assumptions and a complete (and correct) proof?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not include theoretical results.
          \item All the theorems, formulas, and proofs in the paper should be numbered and cross-referenced.
          \item All assumptions should be clearly stated or referenced in the statement of any theorems.
          \item The proofs can either appear in the main paper or the supplemental material, but if they appear in the supplemental material, the authors are encouraged to provide a short proof sketch to provide intuition.
          \item Inversely, any informal proof provided in the core of the paper should be complemented by formal proofs provided in appendix or supplemental material.
          \item Theorems and Lemmas that the proof relies upon should be properly referenced.
        \end{itemize}

  \item {\bf Experimental Result Reproducibility}
  \item[] Question: Does the paper fully disclose all the information needed to reproduce the main experimental results of the paper to the extent that it affects the main claims and/or conclusions of the paper (regardless of whether the code and data are provided or not)?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not include experiments.
          \item If the paper includes experiments, a No answer to this question will not be perceived well by the reviewers: Making the paper reproducible is important, regardless of whether the code and data are provided or not.
          \item If the contribution is a dataset and/or model, the authors should describe the steps taken to make their results reproducible or verifiable.
          \item Depending on the contribution, reproducibility can be accomplished in various ways. For example, if the contribution is a novel architecture, describing the architecture fully might suffice, or if the contribution is a specific model and empirical evaluation, it may be necessary to either make it possible for others to replicate the model with the same dataset, or provide access to the model. In general. releasing code and data is often one good way to accomplish this, but reproducibility can also be provided via detailed instructions for how to replicate the results, access to a hosted model (e.g., in the case of a large language model), releasing of a model checkpoint, or other means that are appropriate to the research performed.
          \item While NeurIPS does not require releasing code, the conference does require all submissions to provide some reasonable avenue for reproducibility, which may depend on the nature of the contribution. For example
                \begin{enumerate}
                  \item If the contribution is primarily a new algorithm, the paper should make it clear how to reproduce that algorithm.
                  \item If the contribution is primarily a new model architecture, the paper should describe the architecture clearly and fully.
                  \item If the contribution is a new model (e.g., a large language model), then there should either be a way to access this model for reproducing the results or a way to reproduce the model (e.g., with an open-source dataset or instructions for how to construct the dataset).
                  \item We recognize that reproducibility may be tricky in some cases, in which case authors are welcome to describe the particular way they provide for reproducibility. In the case of closed-source models, it may be that access to the model is limited in some way (e.g., to registered users), but it should be possible for other researchers to have some path to reproducing or verifying the results.
                \end{enumerate}
        \end{itemize}

  \item {\bf Open access to data and code}
  \item[] Question: Does the paper provide open access to the data and code, with sufficient instructions to faithfully reproduce the main experimental results, as described in supplemental material?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that paper does not include experiments requiring code.
          \item Please see the NeurIPS code and data submission guidelines (\url{https://nips.cc/public/guides/CodeSubmissionPolicy}) for more details.
          \item While we encourage the release of code and data, we understand that this might not be possible, so “No” is an acceptable answer. Papers cannot be rejected simply for not including code, unless this is central to the contribution (e.g., for a new open-source benchmark).
          \item The instructions should contain the exact command and environment needed to run to reproduce the results. See the NeurIPS code and data submission guidelines (\url{https://nips.cc/public/guides/CodeSubmissionPolicy}) for more details.
          \item The authors should provide instructions on data access and preparation, including how to access the raw data, preprocessed data, intermediate data, and generated data, etc.
          \item The authors should provide scripts to reproduce all experimental results for the new proposed method and baselines. If only a subset of experiments are reproducible, they should state which ones are omitted from the script and why.
          \item At submission time, to preserve anonymity, the authors should release anonymized versions (if applicable).
          \item Providing as much information as possible in supplemental material (appended to the paper) is recommended, but including URLs to data and code is permitted.
        \end{itemize}

  \item {\bf Experimental Setting/Details}
  \item[] Question: Does the paper specify all the training and test details (e.g., data splits, hyperparameters, how they were chosen, type of optimizer, etc.) necessary to understand the results?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not include experiments.
          \item The experimental setting should be presented in the core of the paper to a level of detail that is necessary to appreciate the results and make sense of them.
          \item The full details can be provided either with the code, in appendix, or as supplemental material.
        \end{itemize}

  \item {\bf Experiment Statistical Significance}
  \item[] Question: Does the paper report error bars suitably and correctly defined or other appropriate information about the statistical significance of the experiments?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not include experiments.
          \item The authors should answer "Yes" if the results are accompanied by error bars, confidence intervals, or statistical significance tests, at least for the experiments that support the main claims of the paper.
          \item The factors of variability that the error bars are capturing should be clearly stated (for example, train/test split, initialization, random drawing of some parameter, or overall run with given experimental conditions).
          \item The method for calculating the error bars should be explained (closed form formula, call to a library function, bootstrap, etc.)
          \item The assumptions made should be given (e.g., Normally distributed errors).
          \item It should be clear whether the error bar is the standard deviation or the standard error of the mean.
          \item It is OK to report 1-sigma error bars, but one should state it. The authors should preferably report a 2-sigma error bar than state that they have a 96\% CI, if the hypothesis of Normality of errors is not verified.
          \item For asymmetric distributions, the authors should be careful not to show in tables or figures symmetric error bars that would yield results that are out of range (e.g. negative error rates).
          \item If error bars are reported in tables or plots, The authors should explain in the text how they were calculated and reference the corresponding figures or tables in the text.
        \end{itemize}

  \item {\bf Experiments Compute Resources}
  \item[] Question: For each experiment, does the paper provide sufficient information on the computer resources (type of compute workers, memory, time of execution) needed to reproduce the experiments?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not include experiments.
          \item The paper should indicate the type of compute workers CPU or GPU, internal cluster, or cloud provider, including relevant memory and storage.
          \item The paper should provide the amount of compute required for each of the individual experimental runs as well as estimate the total compute.
          \item The paper should disclose whether the full research project required more compute than the experiments reported in the paper (e.g., preliminary or failed experiments that didn't make it into the paper).
        \end{itemize}

  \item {\bf Code Of Ethics}
  \item[] Question: Does the research conducted in the paper conform, in every respect, with the NeurIPS Code of Ethics \url{https://neurips.cc/public/EthicsGuidelines}?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the authors have not reviewed the NeurIPS Code of Ethics.
          \item If the authors answer No, they should explain the special circumstances that require a deviation from the Code of Ethics.
          \item The authors should make sure to preserve anonymity (e.g., if there is a special consideration due to laws or regulations in their jurisdiction).
        \end{itemize}

  \item {\bf Broader Impacts}
  \item[] Question: Does the paper discuss both potential positive societal impacts and negative societal impacts of the work performed?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that there is no societal impact of the work performed.
          \item If the authors answer NA or No, they should explain why their work has no societal impact or why the paper does not address societal impact.
          \item Examples of negative societal impacts include potential malicious or unintended uses (e.g., disinformation, generating fake profiles, surveillance), fairness considerations (e.g., deployment of technologies that could make decisions that unfairly impact specific groups), privacy considerations, and security considerations.
          \item The conference expects that many papers will be foundational research and not tied to particular applications, let alone deployments. However, if there is a direct path to any negative applications, the authors should point it out. For example, it is legitimate to point out that an improvement in the quality of generative models could be used to generate deepfakes for disinformation. On the other hand, it is not needed to point out that a generic algorithm for optimizing neural networks could enable people to train models that generate Deepfakes faster.
          \item The authors should consider possible harms that could arise when the technology is being used as intended and functioning correctly, harms that could arise when the technology is being used as intended but gives incorrect results, and harms following from (intentional or unintentional) misuse of the technology.
          \item If there are negative societal impacts, the authors could also discuss possible mitigation strategies (e.g., gated release of models, providing defenses in addition to attacks, mechanisms for monitoring misuse, mechanisms to monitor how a system learns from feedback over time, improving the efficiency and accessibility of ML).
        \end{itemize}

  \item {\bf Safeguards}
  \item[] Question: Does the paper describe safeguards that have been put in place for responsible release of data or models that have a high risk for misuse (e.g., pretrained language models, image generators, or scraped datasets)?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper poses no such risks.
          \item Released models that have a high risk for misuse or dual-use should be released with necessary safeguards to allow for controlled use of the model, for example by requiring that users adhere to usage guidelines or restrictions to access the model or implementing safety filters.
          \item Datasets that have been scraped from the Internet could pose safety risks. The authors should describe how they avoided releasing unsafe images.
          \item We recognize that providing effective safeguards is challenging, and many papers do not require this, but we encourage authors to take this into account and make a best faith effort.
        \end{itemize}

  \item {\bf Licenses for existing assets}
  \item[] Question: Are the creators or original owners of assets (e.g., code, data, models), used in the paper, properly credited and are the license and terms of use explicitly mentioned and properly respected?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not use existing assets.
          \item The authors should cite the original paper that produced the code package or dataset.
          \item The authors should state which version of the asset is used and, if possible, include a URL.
          \item The name of the license (e.g., CC-BY 4.0) should be included for each asset.
          \item For scraped data from a particular source (e.g., website), the copyright and terms of service of that source should be provided.
          \item If assets are released, the license, copyright information, and terms of use in the package should be provided. For popular datasets, \url{paperswithcode.com/datasets} has curated licenses for some datasets. Their licensing guide can help determine the license of a dataset.
          \item For existing datasets that are re-packaged, both the original license and the license of the derived asset (if it has changed) should be provided.
          \item If this information is not available online, the authors are encouraged to reach out to the asset's creators.
        \end{itemize}

  \item {\bf New Assets}
  \item[] Question: Are new assets introduced in the paper well documented and is the documentation provided alongside the assets?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not release new assets.
          \item Researchers should communicate the details of the dataset/code/model as part of their submissions via structured templates. This includes details about training, license, limitations, etc.
          \item The paper should discuss whether and how consent was obtained from people whose asset is used.
          \item At submission time, remember to anonymize your assets (if applicable). You can either create an anonymized URL or include an anonymized zip file.
        \end{itemize}

  \item {\bf Crowdsourcing and Research with Human Subjects}
  \item[] Question: For crowdsourcing experiments and research with human subjects, does the paper include the full text of instructions given to participants and screenshots, if applicable, as well as details about compensation (if any)?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not involve crowdsourcing nor research with human subjects.
          \item Including this information in the supplemental material is fine, but if the main contribution of the paper involves human subjects, then as much detail as possible should be included in the main paper.
          \item According to the NeurIPS Code of Ethics, workers involved in data collection, curation, or other labor should be paid at least the minimum wage in the country of the data collector.
        \end{itemize}

  \item {\bf Institutional Review Board (IRB) Approvals or Equivalent for Research with Human Subjects}
  \item[] Question: Does the paper describe potential risks incurred by study participants, whether such risks were disclosed to the subjects, and whether Institutional Review Board (IRB) approvals (or an equivalent approval/review based on the requirements of your country or institution) were obtained?
  \item[] Answer: \answerTODO{} % Replace by \answerYes{}, \answerNo{}, or \answerNA{}.
  \item[] Justification: \justificationTODO{}
  \item[] Guidelines:
        \begin{itemize}
          \item The answer NA means that the paper does not involve crowdsourcing nor research with human subjects.
          \item Depending on the country in which research is conducted, IRB approval (or equivalent) may be required for any human subjects research. If you obtained IRB approval, you should clearly state this in the paper.
          \item We recognize that the procedures for this may vary significantly between institutions and locations, and we expect authors to adhere to the NeurIPS Code of Ethics and the guidelines for their institution.
          \item For initial submissions, do not include any information that would break anonymity (if applicable), such as the institution conducting the review.
        \end{itemize}

\end{enumerate}

\end{document}

### 2024-07-29 paper/paper.tex
Add title and authors

@@
 % ready for submission
-\usepackage{neurips_2024}
+\usepackage[preprint]{neurips_2024}
 
@@
 
-\title{Formatting Instructions For NeurIPS 2024}
+\title{Hyperbolic SMC: a Variational Proposal for Sequential Monte Carlo Phylogenetic Inference}
 
@@
 \author{%
-  David S.~Hippocampus\thanks{Use footnote for providing further information
-    about author (webpage, alternative address)---\emph{not} for acknowledging
-    funding agencies.} \\
-  Department of Computer Science\\
-  Cranberry-Lemon University\\
-  Pittsburgh, PA 15213 \\
-  \texttt{hippo@cs.cranberry-lemon.edu} \\
-  % examples of more authors
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \AND
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
-  % \And
-  % Coauthor \\
-  % Affiliation \\
-  % Address \\
-  % \texttt{email} \\
+  Alex Chen \\
+  Department of Computer Science \\
+  Columbia University \\
+  \texttt{azc2110@columbia.edu} \\
+  \And
+  Antonio Khalil Moretti \\
+  Department of Computer Science \\
+  Columbia University \\
+  \texttt{amoretti@cs.columbia.edu} \\
 }

### 2024-07-29 paper/paper.tex
Write abstract

@@
 \begin{abstract}
-  The abstract paragraph should be indented \nicefrac{1}{2}~inch (3~picas) on
-  both the left- and right-hand margins. Use 10~point type, with a vertical
-  spacing (leading) of 11~points.  The word \textbf{Abstract} must be centered,
-  bold, and in point size 12. Two line spaces precede the abstract. The abstract
-  must be limited to one paragraph.
+  Variational combinatorial sequential Monte Carlo (VCSMC) provides a robust and flexible framework for learning distributions over phylogenetic state space.
+In the original VCSMC implementation, learnable branch length parameters are linked to each coalescent event.
+We propose a new variational proposal distribution in which the tips of the phylogenetic tree are embedded in a hyperbolic latent space that captures the evolutionary distance between biological sequences.
+We show that our proposal parameters can be efficiently optimized to maximize the evidence lower bound (ELBO) using stochastic gradient descent.
+Our method, Hyperbolic SMC, produces likelihood estimates that are competitive with state-of-the-art combinatorial sequential Monte Carlo (CSMC) methods.
+Hyperbolic SMC produces tight likelihood estimates on even very large datasets, requires no preliminary tree search, and is highly parallelized to take advantage of modern GPUs.
 \end{abstract}

### 2024-07-29 paper/paper.tex
Add content

@@
 \usepackage{xcolor}         % colors
+
+% additional packages
+\usepackage{siunitx}
+\usepackage{physics}
+
+% additional declarations
+\DeclareMathOperator{\arcosh}{arcosh}
 
@@
 \end{abstract}
+
+\section{Introduction}
+
+\section{Background}
+
+\section{The Hyperbolic SMC proposal}
+
+Consider constructing a binary phylogenetic tree over $N$ biological sequences.
+In the original VCSMC proposal, $2(N-1)$ parameters capture the distributions of branches emanating from the $N-1$ internal nodes.
+A key drawback is that the parameters are identified by coalescent event order, which means the role of each parameter is sensitive to the random order in which nodes are coalesced.
+
+The central idea behind Hyperbolic SMC is to parameterize the branch length distributions by embedding the biological sequences into a hyperbolic latent space, such that distances in this space correspond roughly to evolutionary distances between sequences.
+The Hyperbolic SMC proposal can be explained in three parts:
+
+\begin{enumerate}
+  \item How are biological sequences (tree tips) embedded into latent space?
+  \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, how are the $N-1$ internal nodes embedded?
+  \item For a given branch, how to the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
+\end{enumerate}
+
+\subsection{Embedding biological sequences}
+
+% TODO cite? (https://en.wikipedia.org/wiki/Poincaré_disk_model#Lines_and_distance)
+Many schemes exist for embedding biological sequences, characterized primarily by the embedding dimensionality and distance metric used.
+Hyperbolic SMC uses the Poincaré ball model, in which embeddings exist in a $D$-dimensional vector space.
+The distance between vectors $u$ and $v$ is given by
+
+\begin{equation}
+  d(u, v) = \arcosh\!\qty(1 + 2 \frac{\norm{u - v}^2}{(1 - \norm{u}^2)(1 - \norm{v}^2)}) \; .
+  \label{eq:hyp-distance}
+  \footnote{The gradient of $\arcosh(\cdot)$ is unbounded as its argument approaches $1$. We solve this by adding a small $\epsilon \approx \num{1e-7}$ to the argument.}
+\end{equation}
+
+In this raw form, every embedding $v$ must satisfy $\norm{v} < 1$.
+However, even if all embeddings satisfy this condition initially, gradient updates may push embeddings outside the unit ball.
+To solve this, we represent ``raw'' embeddings as vectors $v' \in \mathbb{R}^D$ that are then ``projected'' into the unit ball under a transformation that affects only magnitude:
+
+\begin{equation}
+  \norm{v} = \sqrt{1 - \exp(-\norm{v'}^2)} \; .
+\end{equation}
+
+Under this particular transformation, the fraction in \ref{eq:hyp-distance} takes a convenient form that is everywhere positive and continuous with respect to the raw embeddings $u'$ and $v'$:
+
+\begin{equation}
+  d(u, v) = \arcosh\!\qty(1 + 2 \, \norm{u - v}^2 \exp(\norm{u'}^2) \exp(\norm{v'}^2)) \; .
+\end{equation}
+
+% TODO cite (find that paper that projected onto a hypersphere)
+Experimentally, this technique yields better results than simply projecting onto a hypersphere of learnable radius.
+
+% TODO mention that D=2 is sufficient, and the 2D Poincaré disk can be easily visualized
+
+\subsection{Embedding internal nodes}
+
+\subsection{Sampling branch lengths}
+
+\subsection{Nested Hyperbolic SMC}
+
+\subsection{Two-phased training}
+
+\section{Computational optimizations}
+
+\subsection{The hash trick}
+
+\subsection{Float32 precision}
+
+\section{Experiments}
+
+\section{Discussion}
 

### 2024-07-29 paper/paper.tex
Fix typo
Add more content

@@
 \usepackage{physics}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 
+\subsection{Motivation}
+
 Consider constructing a binary phylogenetic tree over $N$ biological sequences.
@@
   \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, how are the $N-1$ internal nodes embedded?
-  \item For a given branch, how to the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
+  \item For a given branch, how do the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
 \end{enumerate}
@@
 
+At each coalescent event, two trees from the forest are selected uniformly to be merged under a new internal node.
+Hyperbolic SMC requires a deterministic function that maps the embeddings of two tree roots to the embedding of the new internal node.
+A reasonable choice is the midpoint of the geodesic between the two roots.
+However, a desirable property is that the internal node falls closer to the origin than either of the two roots.
+We want trees to grow outward so that child nodes can leverage the increased space toward the boundary of the unit ball.
+
+% TODO run an experiment to test different internal embedding functions
+An improved heuristic is to first project the further root node onto the same hypersphere as the closer root node.
+Then, the geodesic midpoint is taken as the new internal node.
+A key requirement of the embedding function is that internal node embeddings are differentiable with respect to the embeddings of their subtree roots.
+This allows all internal embeddings to be optimized indirectly via gradient steps on the tip embeddings.
+
+% TODO write up the math for the geodesic midpoint, which is simpler in the case that the two roots are already on the same hypersphere
+
 \subsection{Sampling branch lengths}
+
+% TODO test log normal vs "squared normal" again
+Ultimately, after selecting two trees to merge at a coalescent event, we must manifest the two branch length distributions.
+Whichever class of distributions is used, the child-parent embedding distances should govern the mean of the distribution.
+The original VCSMC proposal uses exponential distribution.
+However, we adopt the commonly used log-normal distribution.
+
+At each coalescent event, the two branch lengths are sampled independently from log-normal distributions with means equal to the embedding distances and variances equal to a global learnable parameter.
+The reparameterization trick is used to propagate gradients through the sampling process to the embeddings.
+
+\subsection{Implementation}
+
+Algorithm \ref{alg:hyperbolic-smc} outlines how the Hyperbolic SMC proposal fits into the VCSMC framework.
+
+% TODO replace this with Hyperbolic SMC
+\begin{algorithm}
+  \caption{Hyperbolic Sequential Monte Carlo}
+  \label{alg:hyperbolic-smc}
+  \begin{algorithmic}[1]
+    \Require $\mathbf{Y} = \{Y_1, \cdots, Y_M\} \in \Omega^{N \times M}$, $\theta = (Q, \{\lambda_i\}_{i=1}^{|E|})$
+    \State \textbf{Initialization.} $\forall k, s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
+    \For{$r = 1$ to $R = N - 1$}
+    \For{$k = 1$ to $K$}
+    \State \textbf{Resample} $\mathbb{P}(a_{r-1}^k = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l}$
+    \For{$i = 1$ to $L = \binom{N-r}{2}$}
+    \For{$m = 1$ to $M$}
+    \State \textbf{Form look-ahead partial state}
+    \[
+      s_r^{k,m}[i] \sim q(\cdot|s_{r-1}^{a_{r-1}^k})
+    \]
+    \State \textbf{Compute potentials}
+    \[
+      w_r^{k,m}[i] = \frac{\pi(s_r^{k,m}[i]) \cdot \nu^{-}(s_{r-1}^{a_{r-1}^k})}{\pi(s_{r-1}^{a_{r-1}^k}) q(s_r^{k,m}[i] | s_{r-1}^{a_{r-1}^k})}
+    \]
+    \EndFor
+    \EndFor
+    \State \textbf{Extend partial state}
+    \[
+      s_r^k = s_r^{k,J}[I], \quad \mathbb{P}(I = i, J = j) = \frac{w_r^{k,j}[i]}{\sum_{l=1}^L \sum_{m=1}^M w_r^{k,m}[l]}
+    \]
+    \State \textbf{Compute weights}
+    \[
+      w_r^k = \frac{1}{ML} \sum_{i=1}^L \sum_{m=1}^M w_r^{k,m}[i]
+    \]
+    \EndFor
+    \EndFor
+    \Ensure $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
+  \end{algorithmic}
+\end{algorithm}
 

### 2024-07-30 paper/paper.tex
Write up algorithm

@@
 
-At each coalescent event, the two branch lengths are sampled independently from log-normal distributions with means equal to the embedding distances and variances equal to a global learnable parameter.
+At each coalescent event, the two branch lengths are sampled independently from log-normal distributions with medians equal to the embedding distances and variances equal to a global learnable parameter.
 The reparameterization trick is used to propagate gradients through the sampling process to the embeddings.
@@
   \begin{algorithmic}[1]
-    \Require $\mathbf{Y} = \{Y_1, \cdots, Y_M\} \in \Omega^{N \times M}$, $\theta = (Q, \{\lambda_i\}_{i=1}^{|E|})$
-    \State \textbf{Initialization.} $\forall k, s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
-    \For{$r = 1$ to $R = N - 1$}
-    \For{$k = 1$ to $K$}
-    \State \textbf{Resample} $\mathbb{P}(a_{r-1}^k = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l}$
-    \For{$i = 1$ to $L = \binom{N-r}{2}$}
-    \For{$m = 1$ to $M$}
-    \State \textbf{Form look-ahead partial state}
+    \State \textbf{Input:} $\theta = (Q, v_0^{1:N}, \sigma)$
+    \State Initialize $\forall k \in [K], s_0^k \leftarrow \bot, w_0^k \leftarrow 1/K$
+    \State Initialize $\forall k \in [K], \forall n \in [N], v_0^{k,n} \leftarrow v_0^n$
+    \For{$r = 1$ \textbf{to} $R = N - 1$}
+    \For{$k \in [K]$}
+    \State Resample particle
     \[
-      s_r^{k,m}[i] \sim q(\cdot|s_{r-1}^{a_{r-1}^k})
+      \mathbb{P}(i_\text{resample} = i) = \frac{w_{r-1}^i}{\sum_{l=1}^K w_{r-1}^l} \; , \forall \, i \in [K]
     \]
-    \State \textbf{Compute potentials}
+    \State Sample two trees uniformly
     \[
-      w_r^{k,m}[i] = \frac{\pi(s_r^{k,m}[i]) \cdot \nu^{-}(s_{r-1}^{a_{r-1}^k})}{\pi(s_{r-1}^{a_{r-1}^k}) q(s_r^{k,m}[i] | s_{r-1}^{a_{r-1}^k})}
+      \mathbb{P}(t_\text{left} = i, t_\text{right} = j) = \frac{1}{\binom{N-r+1}{2}} , \; \forall \, i < j
+    \]
+    \State Form internal embedding $v_\text{parent}$ from $v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}$
+    \State Compute parent-child distances
+    \[
+      d_\text{left} = d\qty(v_{r-1}^{k,t_\text{left}}, v_\text{parent}) \;,\; d_\text{right} = d\qty(v_{r-1}^{k,t_\text{right}}, v_\text{parent})
+    \]
+    \State Sample branch lengths from log-normal distributions (with reparameterization trick)
+    \[
+      b_\text{left} = \exp(\log(d_\text{left}) + \sigma Z_\text{left}) \;,\; b_\text{right} = \exp(\log(d_\text{right}) + \sigma Z_\text{right})
+    \]
+    \[
+      Z_\text{left}, Z_\text{right} \sim \mathcal{N}(0, 1)
+    \]
+    \State Form new partial state $s_r^k$ by merging trees $t_\text{left}, t_\text{right}$
+    \State Compute proposal probability
+    \[
+      q(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
+    \]
+    \State Compute weight
+    \[
+      w_r^k = \frac{\pi(s_r^k) \cdot \nu^-(s_{r-1}^{i_\text{resample}})}{\pi(s_{r-1}^{i_\text{resample}}) \cdot  q(s_r^k | s_{r-1}^{i_\text{resample}})}
+    \]
+    \State
+    Update tree root embeddings
+    \[
+      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{t_\text{left}}, v_{r-1}^{t_\text{right}}}) \cup \qty{v_\text{parent}}
     \]
@@
     \EndFor
-    \State \textbf{Extend partial state}
-    \[
-      s_r^k = s_r^{k,J}[I], \quad \mathbb{P}(I = i, J = j) = \frac{w_r^{k,j}[i]}{\sum_{l=1}^L \sum_{m=1}^M w_r^{k,m}[l]}
-    \]
-    \State \textbf{Compute weights}
-    \[
-      w_r^k = \frac{1}{ML} \sum_{i=1}^L \sum_{m=1}^M w_r^{k,m}[i]
-    \]
-    \EndFor
-    \EndFor
-    \Ensure $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
+    \State \Return $s_{1:R}^{1:K}, w_{1:R}^{1:K}$
   \end{algorithmic}

### 2024-07-31 paper/paper.tex
Fix algo
Write about algorithm

@@
 
+% TODO cite https://www.stats.ox.ac.uk/~doucet/wang_bouchardcote_doucet_BayesianphylogeneticscombinatorialSMC_JASA2015.pdf
 Algorithm \ref{alg:hyperbolic-smc} outlines how the Hyperbolic SMC proposal fits into the VCSMC framework.
+At each optimization step, the algorithm takes as input $Q$, the (possibly learnable) transition matrix, $v_0^{1:N}$, the learned embeddings of the $N$ tree tips, and $\sigma$, the global branch length variance.
+Following the VCSMC framework, at each step $r$, we resample the $k$ particles, then extend each partial state $s_{r-1}^{i_\text{resample}}$ to $s_r^k$ via the proposal distribution $q(\cdot | s_{r-1}^{i_\text{resample}})$.
+The two chosen trees dictate the new internal embedding $v_\text{parent}$, which sets the branch length distributions.
+$\nu^-(\cdot)$ corrects for the overcounting problem detailed in [Wang et al].
+As in the original VCSMC framework, an unbiased estimate for the marginal likelihood is
+\begin{align*}
+  \hat{\mathcal{Z}}_\text{CSMC} = \prod_{r=1}^R \qty(\frac{1}{K} \sum_{k=1}^K w_r^k) \; .
+\end{align*}
 
@@
     \[
-      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{t_\text{left}}, v_{r-1}^{t_\text{right}}}) \cup \qty{v_\text{parent}}
+      \qty{v_r^{k,1}, \cdots, v_r^{k,N-r}} \leftarrow \qty(\qty{v_{r-1}^{k,1}, \cdots, v_{r-1}^{k,N-r+1}} \setminus \qty{v_{r-1}^{k,t_\text{left}}, v_{r-1}^{k,t_\text{right}}}) \cup \qty{v_\text{parent}}
     \]

### 2024-07-31 paper/paper.tex
Break hyp smc into three tasks

@@
 The central idea behind Hyperbolic SMC is to parameterize the branch length distributions by embedding the biological sequences into a hyperbolic latent space, such that distances in this space correspond roughly to evolutionary distances between sequences.
-The Hyperbolic SMC proposal can be explained in three parts:
+The Hyperbolic SMC proposal can be broken into three main tasks:
 
 \begin{enumerate}
-  \item How are biological sequences (tree tips) embedded into latent space?
-  \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, how are the $N-1$ internal nodes embedded?
-  \item For a given branch, how do the embeddings of the two terminal nodes determine the distribution from which the branch length is sampled?
+  \item Embed biological sequences (tree tips) into latent space.
+  \item Given the embeddings of the $N$ tree tips, and prescribed a tree topology, derive embeddings for the $N-1$ internal nodes.
+  \item Sample branch lengths from a distribution parameterized by the embeddings of the two terminal nodes.
 \end{enumerate}
+
+% TODO explain why hyperbolic space is a good choice for tree embeddings
 

### 2024-07-31 paper/paper.tex
Remove comment
Add two-phased training section

@@
 \usepackage{physics}
+\usepackage{cancel}
 \usepackage{algorithm}
@@
 
-% TODO replace this with Hyperbolic SMC
 \begin{algorithm}
@@
     \[
-      q(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
+      q\qty(s_r^k | s_{r-1}^{i_\text{resample}}) = \mathbb{P}(t_\text{left} = i, t_\text{right} = j) \cdot f_\text{log-normal}(b_\text{left} | d_\text{left}, \sigma) \cdot f_\text{log-normal}(b_\text{right} | d_\text{right}, \sigma)
     \]
@@
     \[
-      w_r^k = \frac{\pi(s_r^k) \cdot \nu^-(s_{r-1}^{i_\text{resample}})}{\pi(s_{r-1}^{i_\text{resample}}) \cdot  q(s_r^k | s_{r-1}^{i_\text{resample}})}
+      w_r^k = \frac{\pi\qty(s_r^k) \cdot \nu^-\qty(s_{r-1}^{i_\text{resample}})}{\pi\qty(s_{r-1}^{i_\text{resample}}) \cdot  q\qty(s_r^k | s_{r-1}^{i_\text{resample}})}
     \]
@@
 \subsection{Two-phased training}
+
+Hyperbolic SMC converges especially fast when the variance in the likelihood estimate is minimized.
+A large share of the variance is due to branch length sampling.
+One can imagine an alternate Hyperbolic SMC proposal where parent-child embedding distance are taken directly as branch lengths, with no sampling.
+We must drop the continuous $f_\text{log-normal}(b_\text{left/right} | d_\text{left/right}, \sigma)$ terms from $q(s_r^k | s_{r-1}^{i_\text{resample}})$, but this introduces a bias in the likelihood estimate.
+To see this effect more clearly, consider the expectation (over topology $\tau$ and branch lengths $\mathcal B$) of the log weights:
+
+\begin{equation}
+  \mathbb{E}_{\tau, \mathcal B \sim q\qty(\cdot | s_{r-1}^{i_\text{resample}})} \qty[\log w_r^k]
+  = \dots - \mathbb{E}_{\tau, \mathcal B} \qty[\log q\qty(\tau | s_{r-1}^{i_\text{resample}})]  - \cancel{\mathbb{E}_{\tau, \mathcal B} \qty[\log q\qty(\mathcal B | \tau, s_{r-1}^{i_\text{resample}})]} \; .
+\end{equation}
+
+In the ``deterministic branch length'' setting, we drop the last term the penalizes the ELBO when the branch length distribution is sharply peaked.
+Namely, we overestimate the ELBO by $O(\log (1/\sigma))$, where $\sigma$ is is the interval over which sampling from $\text{Uniform}(b-\sigma, b+\sigma)$ approximates taking $b$ as the branch length.
+
+In practice, we can form a biased likelihood estimator that---despite not being a valid ELBO---allows us to efficiently optimize the tip embeddings.
+In a two-phased training scheme, we first optimize embeddings while simultaneously exploring topology space under the low-variance, biased likelihood estimator.
+Empirically, the learned embeddings form a solid baseline for sampling under the true ELBO (i.e., with proper branch length distributions).
 

### 2024-07-31 scripts/train/hyp_train_hybrid.py
Before hybrid train log normal experiment

@@ -104,7 +104,7 @@ def hyp_train_hybrid(
         N=N,
         lookahead_merge=False,
         sample_branches=True,
-        static_merge_log_weights=static_merge_log_weights,
+        # static_merge_log_weights=static_merge_log_weights,
     )
 
     vcsmc = VCSMC(

### 2024-07-31 vcsmc/proposals.py

@@ -5,7 +5,6 @@ from torch import Tensor, nn
 
 from .distances import Distance
 from .encoders import DummySequenceEncoder, MergeEncoder, SequenceEncoder
-from .utils.distance_utils import EPSILON
 from .utils.repr_utils import custom_module_repr
 from .utils.vcsmc_utils import ArangeFn, gather_K, gather_K2, hash_forest_K
 
@@ -395,12 +394,9 @@ class EmbeddingProposal(Proposal):
             normal1_KxJ = torch.randn([K, J], device=device)
             normal2_KxJ = torch.randn([K, J], device=device)
 
-            dist1_sqrt_KxJ = torch.sqrt(dist1_KxJ + EPSILON)
-            dist2_sqrt_KxJ = torch.sqrt(dist2_KxJ + EPSILON)
-
             # X ~ N(sqrt(dist), sigma)
-            X1_KxJ = dist1_sqrt_KxJ + sigma * normal1_KxJ
-            X2_KxJ = dist2_sqrt_KxJ + sigma * normal2_KxJ
+            X1_KxJ = dist1_KxJ.sqrt() + sigma * normal1_KxJ
+            X2_KxJ = dist2_KxJ.sqrt() + sigma * normal2_KxJ
 
             # Y = X^2
             branch1_KxJ = X1_KxJ**2
@@ -408,8 +404,8 @@ class EmbeddingProposal(Proposal):
 
             # compute log of the pdf of Y
 
-            X1_distr_KxJ = torch.distributions.Normal(dist1_sqrt_KxJ, sigma)
-            X2_distr_KxJ = torch.distributions.Normal(dist2_sqrt_KxJ, sigma)
+            X1_distr_KxJ = torch.distributions.Normal(dist1_KxJ.sqrt(), sigma)
+            X2_distr_KxJ = torch.distributions.Normal(dist2_KxJ.sqrt(), sigma)
 
             # can show: pdf_Y(y) = (1/2|X|) * [pdf_X(X) + pdf_X(-X)]
             log_branch1_prob_KxJ = (

### 2024-07-31 scripts/figures/hyp_smc_init_mean.py
Explicitly set weights_only=False for torch.load

@@ -50,7 +50,8 @@ def load_log_likelihoods(initial_mean: float):
     run_name = f"Hyp_SMC_init_mean{initial_mean}"
 
     results: TrainResults = torch.load(
-        find_most_recent_path(f"runs/*{run_name}", "results.pt")
+        find_most_recent_path(f"runs/*{run_name}", "results.pt"),
+        weights_only=False,
     )
     return results["log_likelihood_avgs"]
 

### 2024-07-31 scripts/figures/proposal_K.py

@@ -84,7 +84,8 @@ def load_log_likelihoods(
         run_name = f"Hyp_NSMC_K{K}"
 
     results: TrainResults = torch.load(
-        find_most_recent_path(f"runs/*{run_name}", "results.pt")
+        find_most_recent_path(f"runs/*{run_name}", "results.pt"),
+        weights_only=False,
     )
     return results["log_likelihood_avgs"]
 

### 2024-07-31 vcsmc/train.py

@@ -355,7 +355,8 @@ def load_checkpoint(
     def find_best_epoch():
         """Returns the epoch with the highest ZCSMC"""
         results: TrainResults = torch.load(
-            find_most_recent_path(search_dir, "results.pt")
+            find_most_recent_path(search_dir, "results.pt"),
+            weights_only=False,
         )
         # there is no off-by-one error here: say epoch i has the highest
         # pre-update LL, so we want to start at epoch i; in this case,
@@ -370,9 +371,13 @@ def load_checkpoint(
         "checkpoint_*.pt" if start_epoch is None else f"checkpoint_{start_epoch}.pt"
     )
 
-    args: TrainArgs = torch.load(find_most_recent_path(search_dir, "args.pt"))
+    args: TrainArgs = torch.load(
+        find_most_recent_path(search_dir, "args.pt"),
+        weights_only=False,
+    )
     checkpoint: TrainCheckpoint = torch.load(
-        find_most_recent_path(search_dir, checkpoint_glob)
+        find_most_recent_path(search_dir, checkpoint_glob),
+        weights_only=False,
     )
 
     start_epoch = checkpoint["start_epoch"]

### 2024-07-31 vcsmc/utils/train_utils.py

@@ -96,10 +96,12 @@ def filter_runs(filter_fn: Callable[[TrainCheckpoint, TrainArgs], bool]):
 
         try:
             args: TrainArgs = torch.load(
-                find_most_recent_path(checkpoints_dir, "args.pt")
+                find_most_recent_path(checkpoints_dir, "args.pt"),
+                weights_only=False,
             )
             checkpoint: TrainCheckpoint = torch.load(
-                find_most_recent_path(checkpoints_dir, "checkpoint_*.pt")
+                find_most_recent_path(checkpoints_dir, "checkpoint_*.pt"),
+                weights_only=False,
             )
 
             matches = filter_fn(checkpoint, args)

### 2024-07-31 scripts/benchmarks/hyp_smc_benchmark.py
Call wandb.finish on error

@@ -3,7 +3,6 @@ import os
 from datetime import datetime
 
 import wandb
-
 from scripts.train.hyp_train_hybrid import hyp_train_hybrid
 from scripts.utils.estimate_latest_run_ll import estimate_latest_run_ll
 from vcsmc.utils.wandb_utils import WANDB_PROJECT, WandbRunType
@@ -53,6 +52,7 @@ def hyp_smc_benchmark():
                 ll_mean, ll_std_dev = estimate_latest_run_ll(ESTIMATE_LL_SAMPLES)
             except Exception as e:
                 print(f"Error while processing {dataset}: {e}")
+                wandb.finish(exit_code=1)
                 continue
 
             ll_mean = round(ll_mean, 2)

### 2024-07-31 scripts/train/hyp_train_hybrid.py
Revert static_merge_log_weights

@@ -104,7 +104,7 @@ def hyp_train_hybrid(
         N=N,
         lookahead_merge=False,
         sample_branches=True,
-        # static_merge_log_weights=static_merge_log_weights,
+        static_merge_log_weights=static_merge_log_weights,
     )
 
     vcsmc = VCSMC(

### 2024-08-01 vcsmc/q_matrix_decoders.py
Remove unused log_stat_probs_A

@@ -118,7 +118,6 @@ class DenseStationaryQMatrixDecoder(QMatrixDecoder):
         self.t_inf = t_inf
 
         self.log_Q_matrix_AxA = nn.Parameter(torch.zeros(A, A))
-        self.log_stat_probs_A = nn.Parameter(torch.zeros(A))
 
     def Q_matrix_VxSxAxA(
         self, embeddings_VxD: Tensor, site_positions_SxC: Tensor
